---
title: 'Lab: Interaction Terms'
author: "Chris Troeger"
date: "2/9/2022"
output: html_document
---

```{r, echo = F, message = F, warning = F}
set.seed(101)
library(knitr)
library(ggplot2)
library(data.table)
library(scales)
library(boot) # has inv.logit() function
library(lme4)
library(MASS)
library(parameters)
```

# Demo or refresher on interaction terms in R

This lab will give a couple examples of specifying and interpreting interaction terms in R.

### Grab Ghana Bednet data

```{r, echo = F}
# Setting personal working directory
os <-  "pc"

if(os  == "mac"){
  setwd("~/Documents/OneDrive - UW/PhD/GH Teaching Assistant/Data")
} else {
  setwd("C:/Users/ctroeger/OneDrive - UW/PhD/GH Teaching Assistant/Data/")
}

# Read in data
dt <- fread("gha_bednet.csv")

```

Let's say we have a hypothesis that there is an interaction between age and effectiveness of bednet. First we will build a basic model and iterate on it. Let's keep it simple and not consider clustering.

```{r}
dt$row <- 1
# make sex a factor for later
dt[, sex := factor(sex, levels = c(0,1), labels = c("Female","Male"))]
tab_dt <- dt[, lapply(.SD, function(x) sum(x)),
             .SDcols = c("row","outcome","follyr"),
             by = c("cluster","bednet","agegp","sex")]
```

My model is:

$$
Y_i \sim \beta_0 + \beta_1bednet_i + \beta_2sex_i + \beta_3agegroup_i + \epsilon_i
$$

Seems like it!

```{r, warning = F, message = F}
mod0 <- glm(outcome ~ offset(log(follyr)) + bednet + agegp + sex, data = dt, family = poisson)
print_md(model_parameters(mod0, exponentiate = T))
```


Let's consider two ways to specify an interaction term between variables. [Useful link](https://conjugateprior.org/2013/01/formulae-in-r-anova/) 

1. Specifying an interaction with the `colon (:)` icon will return a model with those variables *only* as interaction terms. `glm(y ~ a:b)`

$$
Y_i \sim \beta_{a,b}
$$

2. Specifying an interaction with the `asterisk (*)` icon will return a model with those variables separately and as an interaction term. `glm(y ~ a*b)`

$$
Y_i \sim \beta{a} + \beta_b + \beta_{a,b}
$$



```{r, warning = F, message = F}
mod1 <- glm(outcome ~ offset(log(follyr)) + bednet:agegp + sex, data = dt, family = poisson)
print_md(model_parameters(mod1, exponentiate = T))

mod2 <- glm(outcome ~ offset(log(follyr)) + bednet*agegp + sex, data = dt, family = poisson)
print_md(model_parameters(mod2, exponentiate = T))

# mod2 is the same as writing
# glm(outcome ~ offset(log(follyr)) + bednet + agegp + bednet:agegp + sex, data = dt, family = poisson)
```

Interpreting coefficients from interaction terms can be tricky.

* `intercept` = incidence of malaria when `bednet = 0, agegp = 0, sex = "Female"`
* `bednet` = decrease in risk of malaria when `bednet = 1"`
* `agegp` = decrease in risk of malaria for each 1 unit increase in `agegp` when `bednet = 0`
* `sexMale` = decrease in risk of malaria when `sex = "Male"`
* `bednet:agegp` = increase in risk of malaria for each 1 unit increase in `agegp` when `bednet = 1`

Note that we can add interaction terms to a mixed-effects model in exactly the same way:

```{r, warning = F, message = F}
mod3 <- glmer(outcome ~ offset(log(follyr)) + bednet*agegp + sex + (1 | cluster), data = dt, family = poisson)
print_md(model_parameters(mod3, exponentiate = T))
```

### Compare predictions 

```{r}

pred_dt <- expand.grid(agegp = 0:4,
                       sex = c("Male","Female"),
                       bednet = c(0,1),
                       follyr = 1)

# predict model without interaction
pred_dt$pred_oneway <- predict(mod0, newdata = pred_dt)
# predict model with interaction
pred_dt$pred <- predict(mod2, newdata = pred_dt)

ggplot(pred_dt, aes(x = agegp, y = pred_oneway, lty = factor(sex), col = factor(bednet))) + geom_line() +
  theme_classic() + ggtitle("One-way model", subtitle = "Log-space")

ggplot(pred_dt, aes(x = agegp, y = exp(pred_oneway), lty = factor(sex), col = factor(bednet))) + geom_line() +
  theme_classic() + ggtitle("One-way model", subtitle = "Linear-space")

ggplot(pred_dt, aes(x = agegp, y = pred, lty = factor(sex), col = factor(bednet))) + geom_line() + facet_wrap(~sex) +
  theme_classic() + ggtitle("Interaction model", subtitle = "Log-space")

ggplot(pred_dt, aes(x = agegp, y = exp(pred), lty = factor(sex), col = factor(bednet))) + geom_line() +
  theme_classic() + ggtitle("Interaction model", subtitle = "Linear-space")

ggplot(pred_dt, aes(x = pred_oneway, y = pred)) + geom_point() +
  theme_classic()

```