---
title: "Bootstrap Simulations"
author: "Chris Troeger"
date: "2/24/2022"
output: html_document
---

```{r, warning = F, message = F}
library(data.table)
library(ggplot2)
library(parameters)
library(gridExtra)
library(MASS)
library(viridis)
library(GGally)
library(boot)

set.seed(14)

## Intro to rnorm, mvrnorm
mean_vals <- c(10, 0)
vars <- c(2, 5)

x1 <- rnorm(1000, mean_vals[1], vars[1]^2)
hist(x1)

x2 <- rnorm(1000, mean_vals[2], vars[2]^2)
hist(x2)

ggplot(data.frame(x1, x2), aes(x = x1, y = x2)) + geom_point() +
  theme_classic() + stat_density_2d(aes(fill = ..level..), geom="polygon") + scale_fill_viridis(alpha = 0.7) + 
  guides(fill = "none") + ggtitle("Independent")

covs <- 2
var_cov <- matrix(c(vars[1], covs, covs, vars[2]), nrow = 2)

draws <- mvrnorm(1000, mean_vals, var_cov)
draws <- as.data.frame(draws)

ggplot(draws, aes(x = V1, y = V2)) + theme_classic() + geom_point() +
  stat_density_2d(aes(fill = ..level..), geom="polygon") + scale_fill_viridis(alpha = 0.7) + 
  guides(fill = "none") + ggtitle("Conditional (multivariate)")

## Info
cov(x1, x2)
cor(x1, x2)

cov(draws$V1, draws$V2)
cor(draws$V1, draws$V2)

cv <- cov(draws$V1, draws$V2)
p <- cv / sd(draws$V1) / sd(draws$V2)
p

cv <- cor(draws$V1, draws$V2) * sd(draws$V1) * sd(draws$V2)
cv

## Make a function to show how changes interact
sim_multi2 <- function(mean_vals, var_vals, cov){
  var_cov <- matrix(c(var_vals[1], cov, cov, var_vals[2]), nrow = 2)
  draws <- mvrnorm(1000, mean_vals, var_cov)
  draws <- as.data.frame(draws)
  a <- ggplot(draws, aes(x = V1, y = V2)) + theme_classic() + geom_point() +
    stat_density_2d(aes(fill = ..level..), geom="polygon") + scale_fill_viridis(alpha = 0.7) + 
    guides(fill = "none")
  b <- ggplot(draws, aes(x = V1)) + geom_histogram(col = "black") + theme_classic() +
    xlab("")
  c <- ggplot(draws, aes(x= V2)) + geom_histogram(col = "black") + coord_flip() + theme_classic() + xlab("")
  d <- ggplot() + theme_void()
  
  grid.arrange(b ,d, a, c, ncol = 2, widths = c(4, 1.4),
               heights = c(1.4, 4))
}

sim_multi2(mean_vals, var_vals = c(4, 25), cov = 5)
```


## Application from a model

```{r, echo = F, eval = F}
zt <- fread("C:/Users/ctroeger/OneDrive - UW/PhD/GH Teaching Assistant/Data/zamstar.csv")

t <- zt[, lapply(.SD, function(x) sum(x)),
        .SDcols = c("d","n"),
        by = c("ecf","hh")]
t[, incidence := d / n]
t

mod0 <- glm(d ~ offset(log(n)) + country + hh*ecf, data = zt, family = poisson)

summary(mod0)

pred_zt <- expand.grid(country = c(1,2),
                       #strata_tst = c(0,1),
                       hh = c(0,1),
                       ecf = c(0,1),
                       n = 1000)
pred_zt$name <- paste0(ifelse(pred_zt$country == 1, "Zambia", "South Africa"),
                 #", ", ifelse(pred_zt$strata_tst == 0, "Low risk", "High risk"),
                 ", ", ifelse(pred_zt$hh == 0, "No HH","HH"),
                 ", ", ifelse(pred_zt$ecf == 0, "No ECF","ECF"))

pred_mod <- predict(mod0, newdata = pred_zt, se.fit = T)

pred_zt$prediction <- pred_mod[[1]]
pred_zt$se <- pred_mod[[2]]

# Prediction and SE are in log space
pred_zt$lin_pred <- exp(pred_zt$prediction)
pred_zt$lower <- exp(pred_zt$prediction - pred_zt$se * qnorm(0.975))
pred_zt$upper <- exp(pred_zt$prediction + pred_zt$se * qnorm(0.975))

# Let's convert to an incidence
pred_zt <- data.table(pred_zt)
pred_zt[, c("lin_pred","lower","upper") := 
          lapply(.SD, function(x) x/n),
        .SDcols = c("lin_pred","lower","upper")]


zt <- merge(zt, pred_zt, by = c("country","hh","ecf"))

ggplot(pred_zt, aes(x = name, y = lin_pred, ymin = lower, ymax = upper)) +
  geom_point(size = 3) +
  geom_errorbar(width = 0) +
  coord_flip() +
  geom_point(data = zt, aes(y = d/n.x, size = n.x), col = "purple", alpha = 0.4) +
  guides(size = "none") +
  theme_classic() + xlab("") + ylab("Incidence")

mean_vals <- coef(mod0)
varcov <- vcov(mod0)

draws <- mvrnorm(1000, mean_vals, varcov)
draws <- as.data.frame(draws)

ggpairs(draws)

# Simulate 1000 draws from the distribution
draws <- mvrnorm(1000, mean_vals, varcov)

# What is this object? It is 1000 rows and 8 cols
dim(draws)

# Each column is a coefficient from the model (named in object)
# while each row is a conditional realization of the coefficient
# (depends on itself and other variables in the model)
coef(mod0)[1]
mean(draws[,1])

# What do we do with this? 
# Let's say we want to know if the effect of the intervention is statistically significant in South Africa (country = 2) and in the high risk strata (strata = 1)
# Create our values. Make sure they are ordered the same as the draws
colnames(draws)
my_counterfactual <- data.table(intercept = 1,
                                country = c(1, 1),
                                ecf = c(0, 0),
                                hh = c(0, 1),
                                country_strata = 1,
                                ecf_hh = c(0, 0))
my_counterfactual

intervention <- data.table()
control <- data.table()

for(i in 1:1000){
  control <- rbind(control,
                   draws[i,] * my_counterfactual[1,])
  intervention <- rbind(intervention,
                        draws[i,] * my_counterfactual[2,])
  
}

# Now we can sum up terms to get the predicted value by draw
pred_control <- rowSums(control)
pred_intervention <- rowSums(intervention)

# Find the difference
pred_difference <- pred_control - pred_intervention

ggplot(rbind(data.table(v1 = pred_control)[, type := "No ECF"], 
             data.table(v1 = pred_intervention)[, type := "ECF"]), aes(x = type, y = exp(v1))) + geom_boxplot()

# Does the 95% CI cross 0?
mean_diff <- mean(exp(pred_difference))
lower_diff <- quantile(exp(pred_difference), 0.025)
upper_diff <- quantile(exp(pred_difference), 0.975)
res <- data.table(mean_diff, lower_diff, upper_diff)

res
```

### Ghana data

```{r}
dt <- fread("C:/Users/ctroeger/OneDrive - UW/PhD/GH Teaching Assistant/Data/gha_bednet.csv")
dt[, sex := factor(sex, levels = c(0,1), labels = c("Female","Male"))]

mod2 <- glm(outcome ~ offset(log(follyr)) + bednet*agegp + sex, data = dt, family = poisson)
print_md(model_parameters(mod2, exponentiate = T))
```

Interpreting coefficients from interaction terms can be tricky.

* `intercept` = incidence of malaria when `bednet = 0, agegp = 0, sex = "Female"`
* `bednet` = decrease in risk of malaria when `bednet = 1"`
* `agegp` = decrease in risk of malaria for each 1 unit increase in `agegp` when `bednet = 0`
* `sexMale` = decrease in risk of malaria when `sex = "Male"`
* `bednet:agegp` = increase in risk of malaria for each 1 unit increase in `agegp` when `bednet = 1`

### Compare predictions 

```{r}

pred_dt <- expand.grid(agegp = 0:4,
                       sex = c("Male","Female"),
                       bednet = c(0,1),
                       follyr = 1)

# predict model with interaction
pred_dt$pred <- predict(mod2, newdata = pred_dt)

ggplot(pred_dt, aes(x = agegp, y = pred, col = factor(bednet), lty = sex)) + geom_line() +
  theme_classic() + ggtitle("Interaction model", subtitle = "Log-space")

ggplot(pred_dt, aes(x = agegp, y = exp(pred), col = factor(bednet), lty = sex)) + geom_line() + 
  theme_classic() + ggtitle("Interaction model", subtitle = "Linear-space")

# Predict uncertainty

pred_ci <- predict(mod2, newdata = pred_dt, se.fit = T)

pred_dt$pred_lower <- pred_ci[[1]] - pred_ci[[2]] * qnorm(0.975)
pred_dt$pred_upper <- pred_ci[[1]] + pred_ci[[2]] * qnorm(0.975)

ggplot(pred_dt[pred_dt$sex == "Female",], aes(x = agegp, y = exp(pred))) + geom_line(aes(col = factor(bednet))) + facet_wrap(~sex) + geom_ribbon(aes(ymin = exp(pred_lower), ymax = exp(pred_upper), fill = factor(bednet)), alpha = 0.3) +
  guides(fill = "none") +
  theme_classic() + ggtitle("Interaction model", subtitle = "Log-space")

# So maybe we want to know if the risk of mortality is different when children are in each age group between bednet and non-bednet.
# Let's start with a single example. Is the risk of mortality different in age group 0?
mean_vals <- coef(mod2)
varcov <- vcov(mod2)

# Simulate 1000 draws from the distribution
draws <- mvrnorm(1000, mean_vals, varcov)
draws <- as.data.frame(draws)
ggpairs(draws)

# What is this object? It is 1000 rows and 5 cols
dim(draws)
head(draws)

# Each column is a coefficient from the model (named in object)
# while each row is a conditional realization of the coefficient
# (depends on itself and other variables in the model)
coef(mod2)
colMeans(draws)

# What do we do with this? 
draws <- as.data.table(draws)
colnames(draws)

pred_dt <- data.table(pred_dt)
my_cf1 <- pred_dt[sex == "Female" & agegp == 0]

# Modify to match our coefficients
my_cf1$sexMale <- 0

# Think conceptually about what we want
draw1 <- draws[1,]
realization1 <- draw1$`(Intercept)` +  # always get intercept
  draw1$bednet * my_cf1$bednet + # multiply by bednet (0/1)
  draw1$agegp * my_cf1$agegp + # multiply by agegp (0 in this example)
  draw1$sexMale * my_cf1$sexMale + # multiply by sex (only Male == 0)
  draw1$`bednet:agegp` * my_cf1$bednet * my_cf1$agegp # multiply by bednet and agegp

realization1

# So that is length 2 (number of rows of my_cf1)

r1 <- draws$`(Intercept)` +  # always get intercept
  draws$bednet * my_cf1[1]$bednet + # multiply by bednet (0/1)
  draws$agegp * my_cf1[1]$agegp + # multiply by agegp (0 in this example)
  draws$sexMale * my_cf1[1]$sexMale + # multiply by sex (only Male == 0)
  draws$`bednet:agegp` * my_cf1[1]$bednet * my_cf1[1]$agegp # multiply by bednet and agegp

head(r1)

r2 <- draws$`(Intercept)` +  # always get intercept
  draws$bednet * my_cf1[2]$bednet + # multiply by bednet (0/1)
  draws$agegp * my_cf1[2]$agegp + # multiply by agegp (0 in this example)
  draws$sexMale * my_cf1[2]$sexMale + # multiply by sex (only Male == 0)
  draws$`bednet:agegp` * my_cf1[2]$bednet * my_cf1[2]$agegp # multiply by bednet and agegp

head(r2)

ggplot(rbind(data.table(y = r1)[, type := "No bednet"],
             data.table(y = r2)[, type := "Bednet"]), 
       aes(x = type, y = y)) + geom_boxplot() +
  geom_point(alpha = 0.2) + theme_classic()

# Is this difference statistically significant?
r_diff <- r1 - r2
mean(r_diff)
quantile(r_diff, 0.025)
quantile(r_diff, 0.975)

ggplot(rbind(data.table(y = r1)[, type := "No bednet"],
             data.table(y = r2)[, type := "Bednet"]), 
       aes(x = y, fill = type)) + geom_density(alpha = 0.3) +
  theme_classic()

ggplot(data.table(x = r_diff), aes(x = r_diff)) + geom_density(fill = "gray") +
  theme_classic() + geom_vline(xintercept = 0)

# Yes it is!!
```
