---
title: "Survival Analysis"
author: "Chris Troeger"
date: "2/23/2022"
output: html_document
---

# Demonstration of survival analysis in R

### Brazilian Nutrition cohort data

```{r}
library(survival)
library(data.table)
library(RColorBrewer)
library(ggplot2)
library(GGally)
library(parameters)
library(gridExtra)

COLS <- brewer.pal(3, "Set2")

# Setting personal working directory
os <-  "pc"

if(os  == "mac"){
  setwd("~/Documents/OneDrive - UW/PhD/GH Teaching Assistant/Data")
} else {
  setwd("C:/Users/ctroeger/OneDrive - UW/PhD/GH Teaching Assistant/Data/")
}

# Read in data
dt <- fread("Nutrimal_supp.csv")

head(dt)

```

The `Surv()` function creates a "survival object" which we will use as the outcome (response) variable in a statistical model. `Surv(time, event)` where `time` is the followup time and `event` is if the event occurs for an observation. 

Our data already has this value, time to malaria  

```{r}
ggplot(dt, aes(x = time_to_first_malaria)) + 
  geom_histogram(binwidth = 30, col = "black", fill = "lavender") + theme_classic() + ggtitle("All children")

ggplot(dt[malariasn1 == "sim"], aes(x = time_to_first_malaria)) + 
  geom_histogram(binwidth = 30, col = "black", fill = "skyblue") + theme_classic() + ggtitle("Children with malaria")
```

Children appear to have malaria throughout the follow up year. 

The `survfit()` function creates hazard curves by finding the proportion that has not seen an event at each `time_to_first_malaria` step. I'm going to look at growth trajectory as a predictor of risk of malaria. 

I will use `survfit()` to create two lines, one for each growth assessment (adequate/inadequate). This is not a statistical model, it is simply showing the failure rates for each group. 

Two ways are shown. The first is the way that Bobby illustrated in class which uses the base R `plot()` function. The second is a function from a ggplot helper package `GGally`. 

```{r}
# Survival analysis time_to_first_malaria
# set malaria as the outcome
dt$malaria <- ifelse(dt$malariasn1 == "sim", 1, 0)
dt$anemia_entrance <- ifelse(dt$anemia_entrada == "99", NA,
                             ifelse(dt$anemia_entrada == "sim", 1, 0))
dt <- dt[time_to_first_malaria <= 365]
dt[, growth_trajectory := ifelse(diagvelocidadecres == "adequada",
                                 "Adequate", "Inadequate")]


hcurves <- survfit(Surv(time_to_first_malaria, malaria) ~ growth_trajectory, data = dt)

# plot.survfit() is much more efficient to plot the hazard curves
plot(survfit(Surv(time_to_first_malaria, malaria) ~ growth_trajectory, data = dt),col=COLS[1:2], ylim=c(0, 1))
legend("bottomleft", legend=c("Adequate","Inadequate"), lty=c(1 ,1), inset=0.02,col=COLS[1:2])

# There is also a ggplot function in the package GGally
ggsurv(hcurves) + geom_ribbon(aes(ymin=low,ymax=up,fill=group),alpha=0.3) +
  theme_classic() + guides(fill = "none")
```

### Cox Proportional Hazards

So we can see that there is an apparent difference in malaria risk by growth trajectory. Let's use a statistical survival model to see if the difference is statistically significant.

Our function is `coxph()` which will run a Cox proportional hazards model. We use the `Surv()` object as the response in our formula.

```{r}
cmod <- coxph(Surv(time_to_first_malaria, malaria) ~ growth_trajectory, data = dt)
summary(cmod)

ggsurv(survfit(cmod)) + theme_classic()
```

We can again use `survfit()` to plot our Cox model

```{r}
cox_fit <- survfit(cmod, newdata = data.frame(growth_trajectory = c("Adequate", "Inadequate")))

plot(cox_fit,mark=NA,conf.int=TRUE,col=COLS[1:2],lty=c(1,2), ylim = c(0.5,1))
legend("bottomleft", legend=c("Adequate", "Inadequate"), lty=c(1 ,2), inset=0.02,col=COLS[1:2])
```

If we want more control over plotting, we need to understand the survfit objects better. 

Let's start with thinking about what it does with raw data. 

```{r}
my_data <- dt[, lapply(.SD, function(x) sum(x)),
              .SDcols = c("malaria"),
              by = c("time_to_first_malaria","growth_trajectory")]
my_data <- my_data[order(time_to_first_malaria)]

n_inadequate <- 137
n_adequate <- 64
my_data[, children := ifelse(growth_trajectory == "Inadequate",
                              n_inadequate, n_adequate)]

my_data[, events := cumsum(malaria), by = "growth_trajectory"]
my_data[, survival := 1 - events / children]

a <- ggsurv(hcurves) + ggtitle("Survfit")
b <- ggplot(my_data, aes(x = time_to_first_malaria, y = survival, col = growth_trajectory)) + geom_step() + theme_classic() + ggtitle("Manual")

grid.arrange(a, b, ncol = 2)
```

Now, we need to pull information from the `survfit coxph` object. Remember that `survfit` makes hazard curves. It can do this from a `Surv` formula from a `coxph` object. 

The returned information is in a list. Let's investigate that:

```{r, eval = F}
View(cox_fit)
```

With this information, I can figure out how to extract the values I am interested int. 

```{r}

my_estimate <- data.frame(time = rep(cox_fit$time, ncol(cox_fit$surv)),
                      survival = c(cox_fit$surv[,1], cox_fit$surv[,2]),
                      lower = c(cox_fit$lower[,1], cox_fit$lower[,2]),
                      upper = c(cox_fit$upper[,1], cox_fit$upper[,2]),
                      growth_trajectory = c(rep("Adequate", length(cox_fit$time)),
                                            rep("Inadequate", length(cox_fit$time))))

ggplot(my_estimate, aes(x = time, y = survival, col = growth_trajectory)) +
  geom_step() + geom_ribbon(aes(ymin = lower, ymax = upper, fill = growth_trajectory), alpha = 0.4) + guides(fill = "none") +
  geom_point(data = my_data, aes(x = time_to_first_malaria, y = survival,
                                 col = growth_trajectory), alpha = 0.8) +
  theme_classic()

```

Let's compare with if we used a `glm(family = binomial(link = log))`

```{r}
gmod <- glm(malaria ~ growth_trajectory, data = dt, family = binomial(link = "log"))

print_md(model_parameters(cmod, exponentiate = T))
print_md(model_parameters(gmod, exponentiate = T))
```